---
title: "Biostat 203B Homework 1"
author: Rozeta Avetisyan
subtitle: Due Jan 21 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r, eval=T}
sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with 
Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-
documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at 
GitHub using your UCLA email. You'll get GitHub Pro account for free (unlimited 
public and private repositories).

2. Create a **private** repository `biostat-203b-2022-winter` and add `Hua-Zhou`
and `maschepps` as your collaborators with write permission.

3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two 
branches `main` and `develop`. The `develop` branch will be your main 
playground, the place where you develop solution (code) to homework problems and
write up report. The `main` branch will be your presentation area. Submit your 
homework files (R markdown file `Rmd`, `html` file converted from R markdown, 
all code and extra data sets to reproduce results) in `main` branch.

4. After each homework due date, teaching assistant and instructor will check 
out your main branch for grading. Tag each of your homework submissions with tag
names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That 
means if you tag your `hw1` submission after deadline, penalty points will be 
deducted for late submission.

5. After this course, you can make this repository public and use it to 
demonstrate your skill sets on job market.

## Q2. Data ethics training

This exercise (and later in this course) uses the [MIMIC-IV data]
(https://mimic-iv.mit.edu), a freely accessible critical care database developed
by the MIT Lab for Computational Physiology. Follow the instructions at 
<https://mimic.mit.edu/docs/gettingstarted/> to (1) complete the CITI `Data or 
Specimens Only Research` course and (2) obtain the PhysioNet credential for 
using the MIMIC-IV data. Display the verification links to your completion 
report and completion certificate here. (Hint: The CITI training takes a couple 
hours and the PhysioNet credentialing takes a couple days; do not leave it to 
the last minute.)


## Q3. Linux Shell Commands

1. The `/mnt/mimiciv/1.0` folder on teaching server contains data sets from 
MIMIC-IV. Refer to the documentation <https://mimic.mit.edu/docs/iv/> for 
details of data files.  
    ```{bash}
    ls -l /mnt/mimiciv/1.0
    ```
Please, do **not** put these data files into Git; they are big. Do **not** copy 
them into your directory. Do **not** decompress the gz data files. These create 
unnecessary big files on storage and are not big data friendly practices. Just 
read from the data folder `/mnt/mimiciv/1.0` directly in following exercises. 

    Use Bash commands to answer following questions.

2. Display the contents in the folders `core`, `hosp`, `icu`. Why are these data
files distributed as `.csv.gz` files instead of `.csv` (comma separated values) 
files? Read the page <https://mimic.mit.edu/docs/iv/> to understand what's in 
each folder.

**Solution:**
```{bash}
ls -l /mnt/mimiciv/1.0/core
ls -l /mnt/mimiciv/1.0/hosp
ls -l /mnt/mimiciv/1.0/icu
```

3. Briefly describe what bash commands `zcat`, `zless`, `zmore`, and `zgrep` do.
**Solution:**
* Command `zcat` can be used to view the full contents of a .gz file without 
uncompressing it.
* Command `zless` is helpful to view the contents of a .gz file page by page or
line by line again without uncompressing the file.
* Command `zmore` is the less powerful analog of zless, and can help view the
contents of a .gz file page by page without uncompressing it.
* Command `zgrep` is like grep for a .gz file without uncompressing that file

4. What's the output of following bash script?
    ```{bash, eval=T}
    for datafile in /mnt/mimiciv/1.0/core/*.gz
      do
        ls -l $datafile
      done
    ```

**Solution:**
This is the output of the bash script in Q4 above:  
-rw-r--r--. 1 root root 17208966 Jan  4 21:48 
/mnt/mimiciv/1.0/core/admissions.csv.gz   
-rw-r--r--. 1 root root 2955582 Jan  4 21:48 
/mnt/mimiciv/1.0/core/patients.csv.gz   
-rw-r--r--. 1 root root 53014503 Jan  4 21:48 
/mnt/mimiciv/1.0/core/transfers.csv.gz

Display the number of lines in each data file using a similar loop.  
Using the script: 
    ```{bash, eval=T}
    for datafile in /mnt/mimiciv/1.0/hosp/*.gz
      do
        wc -l $datafile
      done
    ```
we are able to grep files that end with .gz from hosp folder displayed below:  
  

similarly, the command below displays files in icu folder:
    ```{bash, eval=T}
    for datafile in /mnt/mimiciv/1.0/icu/*.gz
      do
        wc -l $datafile
      done
    ```


5. Display the first few lines of `admissions.csv.gz`. How many rows are in this
data file? How many unique patients (identified by `subject_id`) are in this 
data file? (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `sort`, 
`uniq`, `wc`, and so on.)

**Solution:**
    ```{bash, eval=T}
   zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | sort | tail -n +2 | awk -F: '{OFS = ","}{print $1}' | uniq | awk 'END { print NR }' 
    ```


6. What are the possible values taken by each of the variable `admission_type`, 
`admission_location`, `insurance`, and `ethnicity`? Also report the count for 
each unique value of these variables. (Hint: combine Linux commands `zcat`, 
`head`/`tail`, `awk`, `uniq -c`, `wc`, and so on.)

**Solution:**
    ```{bash, eval=F}
   awk -F, '{ print $1 }' /mnt/mimiciv/1.0/core/admissions.csv.gz
   
    ```


## Q4. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane 
Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and 
Darcy, your friend thinks that Darcy was the most mentioned. You, however, are 
certain it was Elizabeth. Obtain the full text of the novel from 
<http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local 
folder. 
    ```{bash, eval=FALSE}
    wget -nc http://www.gutenberg.org/cache/epub/42671/pg42671.txt
    ```
**Solution:**
The `wget -nc` command downloads information from a web page specified. It 
starts with that web page IP address and since the -nc option was specified the 
information from that web site will be downloaded only if a file with the 
 name (after the web site there is info on where to store the 
downloaded information) specified by "/" does not already exist.

Explain what `wget -nc` does. Do **not** put this text file `pg42671.txt` in 
Git. Complete the following loop to tabulate the number of times each of the 
four characters is mentioned using Linux commands.
    ```{bash, eval=FALSE}
    wget -nc http://www.gutenberg.org/cache/epub/42671/pg42671.txt
    for char in Elizabeth Jane Lydia Darcy
    do
      echo $char:
      # some commands here
    done
    ```

2. What's the difference between the following two commands?
    ```{bash eval=FALSE}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash eval=FALSE}
    echo 'hello, world' >> test2.txt
    ```
**Solution:**
Both commands will create a file with "hello, world" initially, since the file 
name specified is not in the directory. However, when the files exist, the 1st
command will override whatever is in the existing file with 'hello, world'.
Meanwhile, the 2nd command will just add more 'hello, world' to the next empty
line in the existing text file.

3. Using your favorite text editor (e.g., `vi`), type the following and save 
the file as `middle.sh`:
    ```{bash eval=FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```

    
Using `chmod` make the file executable by the owner, and run 
    ```{bash eval=FALSE}
    ./middle.sh pg42671.txt 20 5
    ```
Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this 
shell script. Why do we need the first line of the shell script?
--CURRENTLY PROBLEMATIC: KEEP GETTING AN ERROR ABOUT PERMISSION DENIED???
--CHANGED THE PERMISSION OF BOTH MIDDLE.SH AND .TXT FILE STILL SAME ERROR

### Q5. More fun with Linux

Try following commands in Bash and interpret the results: `cal`, `cal 2021`, 
`cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, 
`uptime`, `who am i`, `who`, `w`, `id`, `last | head`, 
`echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.

```{bash eval=T}
    cal
```
The command  `cal` shows a calendar with the current day marked black.

```{bash eval=T}
    cal 2021
```
The command  `cal 2021` showed the entire calendar for 2021
  
```{bash eval=T}
    cal 9 1752
```    

It seems like the command `cal 9 1752` is supposed to show the calendar for 
September of 1752, but the calendar is missing dates 09/3-09/13 which thows off 
the rest of the dates.

```{bash eval=T}
   date
```    
The command  `date` prints the current date and time using UTC system.

```{bash eval=T}
    hostname
```    
The command  `hostname` shows the server you are using.   

```{bash eval=T}
    arch
```    
The command  `arch` prints the hardware name (architecture).

```{bash eval=T}
    uname -a
```    
The command  `uname -a` prints system information in the following order:
kernel, node name, kernel release, kernel version, the machine name, processor,
hardware platform and the operating system.
   
```{bash eval=T}
    uptime 
```    
The command  `uptime` prints the current time (not PST), how long the system has
been running, how many users are currently logged in and the average load.

```{bash eval=T}
    who am i
```    
The command `who am i` does not output anything because do not own this process
in RStudio.

```{bash eval=T}
    who 
```    
The command `who` prints information about users who are logged in in the 
following order: uder login names, terminal, the time of user login, and IP 
addresses

```{bash eval=T}
    w
```    
The command  `w` prints the following information about the users currently 
logged in: their name, the name of their terminal, their IP address, time logged 
in, time the terminal is idle, time used by all processes, time used by the 
user’s current process, the current process.

```{bash eval=T}
    id 
```    
The command `id` prints uid and gid, groups, and security context for the 
(current if not specified) user

```{bash eval=T}
    last | head
```    
The command  `last | head` gets the information about the users logged in and
out after the start of the server, then prints only the top 10 of that list

```{bash eval=T}
   echo {con,pre}{sent,fer}{s,ed}
```    
The command  `echo {con,pre}{sent,fer}{s,ed}`

```{bash eval=T}
    time sleep 5
```    
The command `time sleep 5` counts how long the system took to execute the 
sleep 5 command which in turn made the system "sleep" for 5 seconds. This was 
depicted in the breakdown by real, user and sys.


```{bash eval=T}
    history | tail
```    
The command `history | tail`  is supposed to get the history of all commands 
executed by the current user since the server was up, and prints the last 10.
However, there seem to be process ownership problems which cause no output when
run in RStudio.

