---
title: "Biostat 203B Homework 1"
author: Rozeta Avetisyan
subtitle: Due Jan 21 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r, eval=T}
sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at GitHub using your UCLA email. You'll get GitHub Pro account for free (unlimited public and private repositories).

2. Create a **private** repository `biostat-203b-2022-winter` and add `Hua-Zhou` and `maschepps` as your collaborators with write permission.

3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two branches `main` and `develop`. The `develop` branch will be your main playground, the place where you develop solution (code) to homework problems and write up report. The `main` branch will be your presentation area. Submit your homework files (R markdown file `Rmd`, `html` file converted from R markdown, all code and extra data sets to reproduce results) in `main` branch.

4. After each homework due date, teaching assistant and instructor will check out your main branch for grading. Tag each of your homework submissions with tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That means if you tag your `hw1` submission after deadline, penalty points will be deducted for late submission.

5. After this course, you can make this repository public and use it to demonstrate your skill sets on job market.

## Q2. Data ethics training

This exercise (and later in this course) uses the [MIMIC-IV data](https://mimic-iv.mit.edu), a freely accessible critical care database developed by the MIT Lab for Computational Physiology. Follow the instructions at <https://mimic.mit.edu/docs/gettingstarted/> to (1) complete the CITI `Data or Specimens Only Research` course and (2) obtain the PhysioNet credential for using the MIMIC-IV data. Display the verification links to your completion report and completion certificate here. (Hint: The CITI training takes a couple hours and the PhysioNet credentialing takes a couple days; do not leave it to the last minute.)


## Q3. Linux Shell Commands

1. The `/mnt/mimiciv/1.0` folder on teaching server contains data sets from MIMIC-IV. Refer to the documentation <https://mimic.mit.edu/docs/iv/> for details of data files.  
    ```{bash}
    ls -l /mnt/mimiciv/1.0
    ```
Please, do **not** put these data files into Git; they are big. Do **not** copy them into your directory. Do **not** decompress the gz data files. These create unnecessary big files on storage and are not big data friendly practices. Just read from the data folder `/mnt/mimiciv/1.0` directly in following exercises. 

    Use Bash commands to answer following questions.

2. Display the contents in the folders `core`, `hosp`, `icu`. Why are these data files distributed as `.csv.gz` files instead of `.csv` (comma separated values) files? Read the page <https://mimic.mit.edu/docs/iv/> to understand what's in each folder.

**Solution:**
```{bash}
ls -l /mnt/mimiciv/1.0/core
ls -l /mnt/mimiciv/1.0/hosp
ls -l /mnt/mimiciv/1.0/icu
```

3. Briefly describe what bash commands `zcat`, `zless`, `zmore`, and `zgrep` do.
**Solution:**
* Command `zcat` can be used to view the full contents of a .gz file without 
uncompressing it.
* Command `zless` is helpful to view the contents of a .gz file page by page or
line by line again without uncompressing the file.
* Command `zmore` is the less powerful analog of zless, and can help view the
contents of a .gz file page by page without uncompressing it.
* Command `zgrep` is like grep for a .gz file without uncompressing that file

4. What's the output of following bash script?
    ```{bash, eval=F}
    for datafile in /mnt/mimiciv/1.0/core/*.gz
      do
        ls -l $datafile
      done
    ```
Display the number of lines in each data file using a similar loop.

**Solution:**
This is the output of the bash script in Q4 above:
-rw-r--r--. 1 root root 17208966 Jan  4 21:48 
/mnt/mimiciv/1.0/core/admissions.csv.gz
-rw-r--r--. 1 root root 2955582 Jan  4 21:48 
/mnt/mimiciv/1.0/core/patients.csv.gz
-rw-r--r--. 1 root root 53014503 Jan  4 21:48 
/mnt/mimiciv/1.0/core/transfers.csv.gz

Using the script: 
    ```{bash, eval=F}
    for datafile in /mnt/mimiciv/1.0/hosp/*.gz
      do
        ls -l $datafile
      done
    ```
we are able to grep files that end with .gz from hosp folder displayed below:
-rw-r--r--. 1 root root 430049 Jan  4 21:51 
/mnt/mimiciv/1.0/hosp/d_hcpcs.csv.gz
-rw-r--r--. 1 root root 29531152 Jan  4 21:48 
/mnt/mimiciv/1.0/hosp/diagnoses_icd.csv.gz
-rw-r--r--. 1 root root 863239 Jan  4 21:48 
/mnt/mimiciv/1.0/hosp/d_icd_diagnoses.csv.gz
-rw-r--r--. 1 root root 579998 Jan  4 21:48 
/mnt/mimiciv/1.0/hosp/d_icd_procedures.csv.gz
-rw-r--r--. 1 root root 14898 Jan  4 21:48 
/mnt/mimiciv/1.0/hosp/d_labitems.csv.gz
-rw-r--r--. 1 root root 11684062 Jan  4 21:51 
/mnt/mimiciv/1.0/hosp/drgcodes.csv.gz
-rw-r--r--. 1 root root 515763427 Jan  4 21:51 
/mnt/mimiciv/1.0/hosp/emar.csv.gz
-rw-r--r--. 1 root root 476252563 Jan  4 21:48 
/mnt/mimiciv/1.0/hosp/emar_detail.csv.gz
-rw-r--r--. 1 root root 2098831 Jan  4 21:51 
/mnt/mimiciv/1.0/hosp/hcpcsevents.csv.gz
-rw-r--r--. 1 root root 2091865786 Jan  4 21:50 
/mnt/mimiciv/1.0/hosp/labevents.csv.gz
-rw-r--r--. 1 root root 99133381 Jan  4 21:48 
/mnt/mimiciv/1.0/hosp/microbiologyevents.csv.gz
-rw-r--r--. 1 root root 422874088 Jan  4 21:48 
/mnt/mimiciv/1.0/hosp/pharmacy.csv.gz
-rw-r--r--. 1 root root 501381155 Jan  4 21:51 
/mnt/mimiciv/1.0/hosp/poe.csv.gz
-rw-r--r--. 1 root root 24020923 Jan  4 21:48 
/mnt/mimiciv/1.0/hosp/poe_detail.csv.gz
-rw-r--r--. 1 root root 367041717 Jan  4 21:49 
/mnt/mimiciv/1.0/hosp/prescriptions.csv.gz
-rw-r--r--. 1 root root 7750325 Jan  4 21:48 
/mnt/mimiciv/1.0/hosp/procedures_icd.csv.gz
-rw-r--r--. 1 root root 9565293 Jan  4 21:48 
/mnt/mimiciv/1.0/hosp/services.csv.gz

similarly, the command below displays files in icu folder:
    ```{bash, eval=F}
    for datafile in /mnt/mimiciv/1.0/icu/*.gz
      do
        ls -l $datafile
      done
    ```
-rw-r--r--. 1 root root 2350783547 Jan  4 21:54 
/mnt/mimiciv/1.0/icu/chartevents.csv.gz
-rw-r--r--. 1 root root 43296273 Jan  4 21:52 
/mnt/mimiciv/1.0/icu/datetimeevents.csv.gz
-rw-r--r--. 1 root root 55917 Jan  4 21:52 
/mnt/mimiciv/1.0/icu/d_items.csv.gz
-rw-r--r--. 1 root root 2848628 Jan  4 21:52 
/mnt/mimiciv/1.0/icu/icustays.csv.gz
-rw-r--r--. 1 root root 352443512 Jan  4 21:52 
/mnt/mimiciv/1.0/icu/inputevents.csv.gz
-rw-r--r--. 1 root root 37095672 Jan  4 21:52 
/mnt/mimiciv/1.0/icu/outputevents.csv.gz
-rw-r--r--. 1 root root 20567368 Jan  4 21:52 
/mnt/mimiciv/1.0/icu/procedureevents.csv.gz

5. Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `sort`, `uniq`, `wc`, and so on.)

6. What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq -c`, `wc`, and so on.)

## Q4. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
    ```{bash, eval=FALSE}
    wget -nc http://www.gutenberg.org/cache/epub/42671/pg42671.txt
    ```
Explain what `wget -nc` does. Do **not** put this text file `pg42671.txt` in Git. Complete the following loop to tabulate the number of times each of the four characters is mentioned using Linux commands.
    ```{bash, eval=FALSE}
    wget -nc http://www.gutenberg.org/cache/epub/42671/pg42671.txt
    for char in Elizabeth Jane Lydia Darcy
    do
      echo $char:
      # some commands here
    done
    ```

2. What's the difference between the following two commands?
    ```{bash eval=FALSE}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash eval=FALSE}
    echo 'hello, world' >> test2.txt
    ```

3. Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
    ```{bash eval=FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
Using `chmod` make the file executable by the owner, and run 
    ```{bash eval=FALSE}
    ./middle.sh pg42671.txt 20 5
    ```
Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script?

### Q5. More fun with Linux

Try following commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.
