---
title: "Biostat 203B Homework 2"
subtitle: Due Feb 6 @ 11:59PM
author: Rozeta Avetisyan
output: 
  html_document:
    toc: true
    toc_depth: 4 
---

Display machine information for reproducibility:
```{r}
sessionInfo()
```

```{r setup, message=F}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE)
library(tidyverse)
library(data.table)
library(lubridate)
```

```{r}
os <- sessionInfo()$running
if (str_detect(os, "Linux")) {
  mimic_path <- "/mnt/mimiciv/1.0"
} else if (str_detect(os, "macOS")) {
  mimic_path <- "/Users/huazhou/Documents/Box Sync/MIMIC/mimic-iv-1.0"
}
```

In this exercise, we use tidyverse (ggpot2, dplyr, etc) to explore the [MIMIC-IV](https://mimic.mit.edu/docs/iv/) data introduced in [homework 1](https://ucla-biostat-203b.github.io/2022winter/hw/hw1/hw1.html) and to build a cohort of ICU stays.

```{r}
# tree -s -L 2 /Users/huazhou/Documents/Box\ Sync/MIMIC/mimic-iv-1.0
system(str_c("tree -s -L 2 ", shQuote(mimic_path)), intern = TRUE)
```

## Q1. `read.csv` (base R) vs `read_csv` (tidyverse) vs `fread` (data.table)

There are quite a few utilities in R for reading plain text data files. Let us test the speed of reading a moderate sized compressed csv file, `admissions.csv.gz`, by three programs: `read.csv` in base R, `read_csv` in tidyverse, and `fread` in the popular data.table package. 

Which function is fastest? Is there difference in the (default) parsed data types? (Hint: R function `system.time` measures run times.)

For later questions, we stick to the tidyverse.

###Solution to Q1: 
I used the code below to calculate an average read time for functions `read.csv`, `read_csv` and `fread` from 3 experiments. 
```{r}
#Tibble to store the read times in:
read_times <- tibble(
  ST_baseR = 1:3, 
  ET_baseR = 0, 
  ST_tidyR = 0,
  ET_tidyR = 0,
  ST_fread = 0,
  ET_fread = 0,
  RT_tidy = 0,
  RT_baseR = 0,
  RT_fread = 0
)

#storing read times into appropriate columns of the tibble from above:
for(i in 1:nrow(read_times)){
  read_times$ST_baseR[i] <- Sys.time()
  admissions_baseR <- read.csv("/mnt/mimiciv/1.0/core/admissions.csv.gz")
  read_times$ET_baseR[i] <- Sys.time()
  read_times$RT_baseR[i] = read_times$ET_baseR[i] - read_times$ST_baseR[i]
  read_times$ST_tidyR[i] <- Sys.time()
  admissions_tidyR <- read_csv("/mnt/mimiciv/1.0/core/admissions.csv.gz")
  read_times$ET_tidyR[i] <- Sys.time()
  read_times$RT_tidy[i] = read_times$ET_tidyR[i] - read_times$ST_tidyR[i]
  read_times$ST_fread[i] <- Sys.time()
  admissions_fread <- read_csv("/mnt/mimiciv/1.0/core/admissions.csv.gz")
  read_times$ET_fread[i] <- Sys.time()
  read_times$RT_fread[i] = read_times$ET_fread[i] - read_times$ST_fread[i]
}

#getting information about the data types:
str(admissions_baseR)
str(admissions_tidyR)
str(admissions_fread)

#removing data sets not used in later questions:
rm(admissions_baseR)
rm(admissions_fread)
```
    The average read time for `read.csv`, `read_csv` and `fread`, respectively, were `r format(mean(read_times$RT_baseR), digits = 2)`,  
`r format(mean(read_times$RT_tidy), digits = 2)`,  and `r format(mean(read_times$RT_fread), digits = 2)`. On average, tidyverse `read_csv` was faster than the other methods.   
    `read.csv` divides data types to character (including dates) or numeric, turning character variables into factors. Meanwhile, 
    both `read_csv` and `fread` include more detailed information about the data types (i.e. separating dates into POSIX type, which
    makes working with dates easier) and they include attributes of the data types, such as specifying if numeric values are integers 
    or doubles, etc. 


## Q2. ICU stays

`icustays.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/icustays/>) contains data about Intensive Care Units (ICU) stays. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/icustays.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

###Solution to Q2 Part 1: 
1. Import `icustatys.csv.gz` as a tibble `icustays_tble`. 
```{r}
#reading in the data set
icustays_tble <- read_csv("/mnt/mimiciv/1.0/icu/icustays.csv.gz")

#converting it to a tibble
icustays_tble <- as_tibble(icustays_tble)
```


###Solution to Q2 Part 2: 
2. How many unique `subject_id`? Can a `subject_id` have multiple ICU stays? 
```{r}
#practice small dataset
prctc <- icustays_tble %>%
           filter(los>28 & los<30) %>%
           select(subject_id, hadm_id, stay_id)

prctc %>%
  n_distinct(.$subject_id)

n_distinct(prctc$subject_id)

#reading in the data set
unique_ID <- icustays_tble %>%
                  n_distinct(.$subject_id)
```
The number of unique `subject_id`s is `r format(mean(unique_ID), digits = 2)`.
Yes, some `subject_id`s have more than 1 `head_id`. 

###Solution to Q2 Part 3: 
3. For each `subject_id`, let's only keep the first ICU stay in the tibble `icustays_tble`.

```{r}
icustays_tble <- icustays_tble %>%
                   arrange(intime) %>%
                   distinct(subject_id, .keep_all = T)

```

## Q3. `admission` data

Information of the patients admitted into hospital is available in `admissions.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/admissions/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/core/admissions.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `admissions.csv.gz` as a tibble `admissions_tble`.
###Solution to Q3 Part 1: 

```{r}
#COMMENT BEFORE KNITTING
#using this when skipping first chunks!
admissions_tidyR <- read_csv("/mnt/mimiciv/1.0/core/admissions.csv.gz")
#converting the data set imported in Q1 to a tibble
admissions_tble <- as_tibble( admissions_tidyR)
```

###Solution to Q3 Part 3: 
2. Let's only keep the admissions that have a match in `icustays_tble` according to `subject_id` and `hadm_id`.
```{r}
admissions_tble <- inner_join(admissions_tble, 
                   icustays_tble, by = c("subject_id", "hadm_id")) 
str(admissions_tble)
```

###Solution to Q3 Part 3: 
3. Summarize the following variables by graphics. 
- admission year  
```{r}
admissions_tble 
admission_year <- year(admissions_tble$admittime)
hist(admission_year)
admissions_tble %>% 
  summarise(
    count = n(),
    average_year = mean(admittime)
  )

```

- admission month  
- admission month day  
- admission week day  
- admission hour (anything unusual?)  


## Q4. `patients` data

Patient information is available in `patients.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/patients/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/core/patients.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

###Solution to Q4 Part 1: 
1. Import `patients.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/core/patients/>) as a tibble `patients_tble` and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).

```{r}
#COMMENT BEFORE KNITTING
#using this when skipping first chunks!
patients_tble <- read_csv("/mnt/mimiciv/1.0/core/patients.csv.gz")
#converting the data set imported in Q1 to a tibble
patients_tble <- as_tibble( patients_tble)
patients_tble <- inner_join(patients_tble, icustays_tble, by = c("subject_id")) 

```


2. Summarize variables `gender` and `anchor_age`, and explain any patterns you see.
```{r}


```



## Q5. Lab results

`labevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/labevents/>) contains all laboratory measurements for patients. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/labevents.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```
`d_labitems.csv.gz` is the dictionary of lab measurements. 
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/d_labitems.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Find how many rows are in `labevents.csv.gz`.

2. We are interested in the lab measurements of creatinine (50912), potassium (50971), sodium (50983), chloride (50902), bicarbonate (50882), hematocrit (51221), white blood cell count (51301), glucose (50931), magnesium (50960), and calcium (50893). Retrieve a subset of `labevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `labevents_tble`. 

    Hint: `labevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` and `lazy` options in the `read_csv` function to reduce the memory burden.

3. Further restrict `labevents_tble` to the first lab measurement during the ICU stay. 
4. Summarize the lab measurements by appropriate numerics and graphics. 

## Q6. Vitals from charted events

`chartevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `itemid` variable indicates a single measurement type in the database. The `value` variable is the value measured for `itemid`. The first 10 lines of `chartevents.csv.gz` are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/chartevents.csv.gz")), 
    " | head"), 
  intern = TRUE
)
```
`d_items.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/d_items/>) is the dictionary for the `itemid` in `chartevents.csv.gz`. 
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/d_items.csv.gz")), 
    " | head"), 
  intern = TRUE
)
```

1. We are interested in the vitals for ICU patients: heart rate (220045), mean non-invasive blood pressure (220181), systolic non-invasive blood pressure (220179), body temperature in Fahrenheit (223761), and respiratory rate (220210). Retrieve a subset of `chartevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `chartevents_tble`.

2. Further restrict `chartevents_tble` to the first vital measurement during the ICU stay. 

3. Summarize these vital measurements by appropriate numerics and graphics. 

## Q7. Putting things together

Let us create a tibble `mimic_icu_cohort` for all ICU stays, where rows are  

- first ICU stay of each unique adult (age at admission > 18)

and columns contain at least following variables  

- all variables in `icustays.csv.gz`  
- all variables in `admission.csv.gz`  
- all variables in `patients.csv.gz`  
- first lab measurements during ICU stay  
- first vital measurements during ICU stay
- an indicator variable `thirty_day_mort` whether the patient died within 30 days of hospital admission (30 day mortality)

## Q8. Exploratory data analysis (EDA)

Summarize following information using appropriate numerics or graphs.

- `thirty_day_mort` vs demographic variables (ethnicity, language, insurance, marital_status, gender, age at hospital admission)

- `thirty_day_mort` vs first lab measurements

- `thirty_day_mort` vs first vital measurements

- `thirty_day_mort` vs first ICU unit

